
---
title: "Pasta Example Analysis: Liu Polo 2020"
author: "Jérôme Salignon"
date: "`r Sys.Date()`"
output:
 md_document:
  variant: gfm
  toc: true
---


# Introduction

This document provides an example analysis using Pasta on dataset **GSE149694** from Liu & Polo, *Nature*, 2020. The study, titled _"Reprogramming roadmap reveals route to human induced trophoblast stem cells"_, is available online at [GEO]((https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE149694)), [Nature](https://www.nature.com/articles/s41586-020-2734-6), and on the [HRPI website](http://hrpi.ddnetbio.com/).

# Data acquisition, age-prediction, and processing

In this section, we download the GEO ExpressionSet, obtain phenotype data with age scores, and perform some data wrangling. The following code executes the complete workflow using the Pasta functions.

## Loading all needed libraries

```{r setup, include=FALSE}
library(pasta)
library(magrittr)
library(data.table)
library(ggplot2)
library(gtools)
library(jsutil)
```

## Getting GEO ExpressionSet and predicted age scores

```{r get-ES, message=FALSE, warning=FALSE}

file_LiuPolo2020 = 'output/ES_LiuPolo2020.rds'
if(!file.exists(file_LiuPolo2020)){
  dir.create('output', showWarnings = F)
  # Download ExpressionSet for GSE149694
  ES <- getting_GEO_ES_for_age_model('GSE149694') %T>% pdim  # Dimensions: 8113 x 8
  saveRDS(ES, file_LiuPolo2020)
}

ES = readRDS(file_LiuPolo2020)
# Obtain phenotype data with age predictions
pdata <- getting_pdata_with_age_scores(ES)
# Subset phenotype data for plotting
pdata1 <- pdata[, c('title', 'PASTA', 'REG', 'CT46')]
```

## Reshaping to long format and cleaning the metadata

Here we reshape the data and extract information such as condition, time, and replicate details from the title.

```{r reshape-data, message=FALSE, warning=FALSE}
dt <- melt(pdata1, id.vars = 'title', variable.name = 'model_type', value.name = 'age_score')
dt[, title := gsub('RNA-seq_', '', title)]
dt[, condition := gsub('-.*', '', title)]
dt[, time := gsub('.*-(.*)_rep.*', '\\1', title)]
dt[, time := factor(time, levels = gtools::mixedsort(unique(dt$time)))]
dt[, rep := gsub('.*rep', '', title)]
dt[, title := NULL]
dt[, time1 := as.integer(time)]
```


## Adding initial time points

The initial time points (Fibroblast, day 3 and 7) is the same between all conditions ('NHSM', '5iLAF', 'Primed', 't2iLGoY', 'RSeT').
missing for all samples. We add this time point to each condition so we can then compute correlation.

```{r load-adjust-data, message=FALSE, warning=FALSE}
dt_pasta <- dt[model_type == 'PASTA']

# Separate fibroblast and non-fibroblast data
dt_fibro <- dt_pasta[condition == 'Fibroblast']
dt_liuPolo <- dt_pasta[condition != 'Fibroblast']

# For each non-fibroblast condition, append a copy of the fibroblast data with the condition overwritten
for(condition1 in unique(dt_liuPolo$condition)){
  dt_fibro1 <- copy(dt_fibro)[, condition := condition1]
  dt_liuPolo <- rbind(dt_liuPolo, dt_fibro1)
}

# Reformat time variables
dt_liuPolo[, time1 := time %>% gsub('D', '', .) %>% gsub('P3', '30', .) %>% 
  gsub('P10', '30', .) %>% as.integer]
# Display unique condition and time mappings
unique(dt_liuPolo[, c('condition', 'time', 'time1')])[order(condition, time1)]
```


# Results

## Computing correlations

We calculate both Pearson and Spearman correlations between predicted age scores and the true time values.

```{r compute-correlations, message=FALSE, warning=FALSE}
dt_cor <- dt_liuPolo[, .(
  PCC = cor(age_score, time1, method = 'pearson'), 
  SCC = cor(age_score, time1, method = 'spearman')
), by = c('model_type', 'condition')][order(SCC)]
print(dt_cor)
```

We can also display overall correlation values.

```{r overall-correlation, message=FALSE, warning=FALSE}
cat("Pearson correlation between time1 and age_score:",
    cor(dt_liuPolo$time1, dt_liuPolo$age_score), "\n")
cat("Spearman correlation between time1 and age_score:",
    cor(dt_liuPolo$time1, dt_liuPolo$age_score, method = 'spearman'), "\n")
```


## Visualization

Finally, we visualize the correlation data and the timecourse of the age predictions.

```{r plot-timecourse, message=FALSE, warning=FALSE}
# Plot predicted age scores over time for the Liu & Polo dataset
p_LiuPolo2020 <- ggplot(dt_liuPolo, aes(x = time, y = age_score)) + 
  geom_point(size = 1, alpha = 0.7) +
  ggtitle('Reprogramming Timecourse') +
  theme_minimal()  # Replace with your custom theme if available

print(p_LiuPolo2020)
```


